# Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
data <- read.table("data\\learning2014.csv", header = T, sep = ",")
head(data)
```

```{r}
str(data)
```

```{r}
dim(data)
```

Here we go again...

```{r}
suppressWarnings(library(ggplot2))
suppressWarnings(library(GGally))
```

```{r}
ggpairs(data, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)),
        upper = list(continuous = wrap("cor", size = 2.5))) 
```

Go on...

```{r}
summary(data$gender)
summary(data$Age)
summary(data$Attitude)
summary(data$deep)
summary(data$stra)
summary(data$surf)
summary(data$points)
```

Go on

```{r}
model1 <- lm(Points ~ Attitude + surf + stra, data = data)
summary(model1)
```

Bla bla bla

```{r}
model2 <- lm(Points ~ Attitude, data = data)
summary(model2)
```

Bla bla bla

```{r}
par(mfrow = c(2,2))
plot(model2, which = c(1, 2, 5))
```