# Dimensionality Reduction Techniques

*What follows is my report for the IODS course exercise 5 that focuses on principal component analysis (PCA) and multiple correspondence analysis (MCA) as dimensionality reduction techniques.*

The data used for this exercise comes from the United Nations Development Programme (see <http://hdr.undp.org/en/content/human-development-index-hdi>) and is a combination of two data sets containing information on the Human Development Index (HDI) and the Gender Inequality Index (GII) for most countries of the world.

```{r message = FALSE, warning = FALSE}
# necessary libraries
library(GGally)
library(corrplot)
library(FactoMineR)
library(dplyr)
library(tidyr)
library(gridExtra)
```

```{r}
# read in previously prepared data
human <- read.table("data\\human.csv", header = TRUE, sep = ",")
```

After some preparation steps, the resulting data set contains observations on 155 countries and the following 8 (mainly numerical) variables:

* *sec_edu_sexratio*: Ratio of female and male population with secondary education
* *lfp_sexratio*: Ratio of female and male labor force participation
* *exp_edu*: Expected years of schooling
* *e0*: Life expectancy at birth
* *gni*: Gross National Income
* *mmr*: Maternal Mortality Ratio
* *abr*: Adolescent birth rate
* *parlia*: Percentage of female representatives in parliament

```{r}
# explore structure and dimensions of the data
str(human)
dim(human)
```

To further explore the data and the relationship between its variables, I prepared visualizations of variable pairs and the correlation matrix, and statistical summaries of the variables.  ...

```{r}
# plots
ggpairs(human, upper = list(continuous = wrap("cor", size = 2.5))) 
corrplot(cor(human))

# statistical summaries
summary(human)
```

```{r}
# PCA (with the SVD method)
pca_human <- prcomp(human)

# principal component's captured variability
summary(pca_human)

# biplot
biplot(pca_human, choices = 1:2, cex = c(0.65, 1), col = c("grey40", "deeppink2"),
       xlab = "PC1 (>99.99%)", ylab = "PC2 (<0.01%)")
```

```{r}
# standardize the data
human_std <- scale(human)

# PCA (with the SVD method)
pca_human_std <- prcomp(human_std)

# principal component's captured variability
summary(pca_human_std)

# biplot
biplot(pca_human_std, choices = 1:2, cex = c(0.5, 0.8), col = c("grey40", "deeppink2"),
       xlim = c(-0.3, 0.3), xlab = "PC1 (53.61%)", ylab = "PC2 (16.24%)")
text(x = 0, y = 15, "Empowerment\nLabor market", cex = 0.7, col = "darkblue")
text(x = 20, y = 1, "Health", cex = 0.7, col = "darkblue")
text(x = -20, y = 3, "Long life", cex = 0.7, col = "darkgreen")
text(x = -18, y = -2, "Knowledge", cex = 0.7, col = "darkgreen")
text(x = -16, y = -4.5, "Standard of living", cex = 0.7, col = "darkgreen")
text(x = 20, y = 10, "GII", col = "darkblue")
text(x = -20, y = 6, "HDI", col = "darkgreen")
```

```{r}
# load data from package FactoMineR
data(tea)

# structure and dimensions of the data
str(tea)
dim(tea)

# filter data to chosen variables
vars <- c("How", "sugar", "where", "age_Q", "sex", "effect.on.health", "frequency")
my_tea <- dplyr::select(tea, one_of(vars))

# rename two variables
names(my_tea)[c(4, 6)] <- c("age", "health.effect")

# visualization of my data
a <- ggplot(data = my_tea, aes(x = How)) + geom_bar(stat = "count")
b <- ggplot(data = my_tea, aes(x = sugar)) + geom_bar(stat = "count")
c <- ggplot(data = my_tea, aes(x = where)) + geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
d <- ggplot(data = my_tea, aes(x = age)) + geom_bar(stat = "count")
e <- ggplot(data = my_tea, aes(x = sex)) + geom_bar(stat = "count")
f <- ggplot(data = my_tea, aes(x = health.effect)) + geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
grid.arrange(a, b, c, d, e, f, nrow = 2)
```

```{r}
# MCA on my data
mca <- MCA(my_tea, graph = TRUE)
summary(mca)
```